---
title: "Differential expression analysis e"
author: "Bo Yu, Si Liu, Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# explicitly set warning level in R, to prevent a warning from being converted to an error
options(warn=1)
```

Differential expression using the bulk RNAseq JAX_RNAseq08 data from the June 2025 release.

See bulk_RNAseq.xlsx for details of the dataset.

For PrS_day_6_nor_ASCL2_ELF5_NR2F2_nor, 20250327_GT25-RobsonP-14-run3, one of the two run_id values, has only one sample. The run_id covariate is ignored for the DE analysis.

## R libraries.

```{r load_libraries, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(stringr)
library(MASS)
library(RColorBrewer)
library(DESeq2)
library(viridis)
library(ggpointdensity)
library(tidyverse)
library(data.table)
library(pheatmap)
library(ComplexHeatmap)
library(UpSetR)
library(readxl)
library(ggsci)
library(ggpointdensity)
library(ggprism)
theme_set(theme_classic(base_size = 10))

personal_path = "/fh/working/sun_w/sshen/MorPhiC"
path = "/fh/fast/sun_w/MorPhiC/data/MorPhiC_exchange_experiment/"
dat_path = file.path(path, "DRACC-processed/JAX_exchange_experiment_DRACC_processed_August2025/Tables")
```

## Read in meta data

```{r fig.width = 3, fig.height=3}
meta = fread(file.path(personal_path, "JAX_meta_data.tsv"), data.table = FALSE)
dim(meta)
meta[1:2, ]
names(meta)
names(meta)[apply(meta, 2, \(x) any(is.na(x)))]

table(meta$clonal.label)
table(table(meta$clonal.label))
table(table(meta$label))

meta$timepoint = as.factor(meta$timepoint_value)
```

## Read in gene count data

```{r}
cts = fread(file.path(dat_path, "genesCounts.csv"), data.table = FALSE)
dim(cts)
cts[1:2, c(1:2, (ncol(cts) - 1) : ncol(cts))]
# Align counts with meta data
stopifnot(
  setequal(meta$file_id, names(cts)[-1]),
  !any(duplicated(meta$file_id)),
  !any(duplicated(names(cts)[-1]))
)
meta = meta[match(names(cts)[-1], meta$file_id), ]
# To matrix
cts_dat = data.matrix(cts[, -1])
rownames(cts_dat) = cts$Name
```

### Exclude genes without annotation

```{r}
# Read in gene annotation
gene_anno = fread("/fh/fast/sun_w/sshen/MorPhiC/gencode_v44_primary_assembly_info.tsv")
dim(gene_anno)
gene_anno[1:2, ]
# Make sure all KO genes have unique ensembl gene id
gene_anno |> 
  filter(hgnc_symbol %in% setdiff(meta$ko_gene, 'WT')) |>
  arrange(hgnc_symbol)
# Exclude genes without annotation
table(rownames(cts_dat) %in% gene_anno$ensembl_gene_id)
cts = cts[cts$Name %in% gene_anno$ensembl_gene_id, ]
cts_dat = cts_dat[rownames(cts_dat) %in% gene_anno$ensembl_gene_id, ]
```

### Discard genes whose expression is 0 in more than 75% of samples

```{r}
gene_info = data.frame(
  apply(cts_dat, 1, min), 
  apply(cts_dat, 1, median), 
  apply(cts_dat, 1, \(x) quantile(x, probs = 0.75)), 
  apply(cts_dat, 1, max)
)
gene_info[1:2, ]
names(gene_info) = c("min", "median", "q75", "max")
summary(gene_info)
table(gene_info$q75 > 0)

# Subset genes
w2kp = which(gene_info$q75 > 0)
cts_dat = cts_dat[w2kp, ]
gene_info = gene_info[w2kp, ]
```

### Check read depth vs. gene expression quantile

```{r, fig.width = 4, fig.height=3}
meta$read_depth = colSums(cts_dat)/1e6
meta$q75 = apply(cts_dat, 2, quantile, probs = 0.75)

my_colors <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", 
               "#FFFF33", "#A65628", "#F781BF", "#999999", "#66C2A5", "#FC8D62")
ggplot(meta, aes(x=read_depth, y=q75, color=ko_strategy, shape = ko_gene)) +
  geom_point(size=2, alpha=0.7) + ggtitle("Gene counts") + 
  scale_shape_manual(values = c(7, 16)) + 
  xlab("read depth (million)") + ylab("75 percentile of gene expression") + 
  labs(color = "KO type", shape = "KO gene") + 
  # scale_color_brewer(palette = "Set1")
  scale_color_manual(values = my_colors)
```

```{r fig.width = 5.2, fig.height=3}
unique(meta$run_id)
meta$run_id_short = meta$run_id
```


```{r}
# PCA of all samples
q75 = apply(cts_dat, 1, quantile, probs=0.75)
cts_dat_n = t(cts_dat[q75 > 0,])
cts_dat_n = log(median(meta$q75)*cts_dat_n/meta$q75 + 1)
dim(cts_dat_n)

sample_pca = prcomp(cts_dat_n, center = TRUE, scale. = TRUE)
names(sample_pca)
pc_scores = sample_pca$x
eigen_vals = sample_pca$sdev^2
eigen_vals[1:5]/sum(eigen_vals)
pvs = round(eigen_vals[1:5]/sum(eigen_vals)*100,1)
pvs

PC_df = data.frame(pc_scores)
PC_df = cbind(PC_df, meta)

gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=ko_gene, color=ko_strategy)) +
  geom_point(size=2.5, alpha=0.7) + 
  scale_color_manual(values = my_colors) +
  scale_shape_prism() + 
  labs(color="KO type", shape ="KO gene") + 
  xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
  ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
  ggtitle("Wild type samples") + 
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  ) +
  theme(
    legend.margin = margin(0, 0, 0, 0), 
    legend.box.just = "left", 
    legend.direction = "vertical" 
  )

print(gPC)
```


## Check PCA result of WT samples

```{r fig.width = 6, fig.height=4.5}
sample2kp = which(meta$ko_gene == "WT")
cts_dat_m = cts_dat[,sample2kp]
meta_m    = meta[sample2kp,]
summary(meta_m$q75)
dim(cts_dat_m)

stopifnot(all(meta_m$file_id == colnames(cts_dat_m)))
table(meta_m$file_id == colnames(cts_dat_m), useNA="ifany")

q75 = apply(cts_dat_m, 1, quantile, probs=0.75)
cts_dat_n = t(cts_dat_m[q75 > 0,])
cts_dat_n = log(median(meta_m$q75)*cts_dat_n/meta_m$q75 + 1)
dim(cts_dat_n)

sample_pca = prcomp(cts_dat_n, center = TRUE, scale. = TRUE)
names(sample_pca)
pc_scores = sample_pca$x
eigen_vals = sample_pca$sdev^2
eigen_vals[1:5]/sum(eigen_vals)
pvs = round(eigen_vals[1:5]/sum(eigen_vals)*100,1)
pvs

PC_df = data.frame(pc_scores)
PC_df = cbind(PC_df, meta_m)

gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=ko_gene, color=ko_strategy)) +
  geom_point(size=2.5, alpha=0.7) + scale_color_brewer(palette = "Dark2") +
  scale_shape_prism() + 
  labs(color="KO type", shape ="KO gene") + 
  xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
  ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
  ggtitle("Wild type samples") + 
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  ) +
  theme(
    legend.margin = margin(0, 0, 0, 0), 
    legend.box.just = "left", 
    legend.direction = "vertical" 
  )

print(gPC)
```

Run DE analysis for each gene-condition level separately. 
They natually separate all batches, except for one sample from 20250327_GT25-RobsonP-14-run3, which presents no batch effect based on the PC plot. 

```{r}
# table(meta$gene_condition_group, meta$run_id, meta$model_system, meta$timepoint)
```

For the output files, create two versions, one with direct information, one with padj set to NA for the genes when the number of 0 per test is >=4. 

In more details, for both versions of the outputs, make it one file per knockout gene per DE group (DE group can be model system, model system + condition, model system + run_id, model system + day, etc.). 

The first version of the output files contains 
    
    gene id, gene symbol

and for each knockout strategy:

    baseMean, log2FoldChange, lfcSE, stat, pvalue, padj
    
The second version of the output files contains  

    gene id, gene symbol, chr, strand, position

and for each knockout strategy:

    log2FoldChange, pvalue, padj (set to be NA if the number of 0 per test >= 4).

In addition, save out a separate file of the sample size for each comparison. This needs to be by DE group. 

### Prepare output directories for DE results

```{r message=FALSE, warning=FALSE}
# 
# release_name = "2025_09_JAX_RNAseq09"
# 
# output_dir = file.path("results", release_name)
# raw_output_dir = file.path(output_dir, "raw")
# processed_output_dir = file.path(output_dir, "processed")
# 
# if (!file.exists(raw_output_dir)){
#   dir.create(raw_output_dir, recursive = TRUE)
# }
# 
# if (!file.exists(processed_output_dir)){
#   dir.create(processed_output_dir, recursive = TRUE)
# }

```

### DE analysis for all genes
```{r fig.height=3.2, fig.width=4.8, message=FALSE, warning=FALSE}
meta$model_condition = stringr::str_sub(meta$model_system, start = 1, end = -2)

df_size = NULL
model1 = c("JAX_C01_Rep", "JAX_E04_Rep", "MSK_Clone1_Rep", "MSK_Clone2_Rep", "JAX_MSK_WT")
sample2kp = which(meta$model_condition %in% model1)

cts_dat_m = cts_dat[,sample2kp]
meta_m    = meta[sample2kp,]

table(meta_m$ko_strategy)
stopifnot(all(meta_m$file_id == colnames(cts_dat_m)))
  
## Generate sample information matrix
cols2kp = c("model_condition", "ko_strategy", "ko_gene", 
            "run_id", "run_id_short", "timepoint", "q75")
sample_info = meta_m[,cols2kp]
colnames(sample_info)[which(colnames(sample_info)=="timepoint")] = "day"

dim(sample_info)
sample_info[1:2,]
  
# do two steps of filtering
# first, filter based on q75 of each gene
# second, filter based on whether each gene is expressed in at least 75% of the samples

dg_q75 = apply(cts_dat_m, 1, quantile, probs=0.75)
cts_dat_m = cts_dat_m[dg_q75 > 0, ]

print(sprintf("After filtering by gene expression q75, the number of genes reduces from %d to %d", 
              length(dg_q75), nrow(cts_dat_m)))

# update the q75 based on only genes that will be used in the DE analysis
sample_info$q75 = apply(cts_dat_m, 2, quantile, probs = 0.75)
sample_info$log_q75 = log(sample_info$q75)

model1 = c("JAX_C01_Rep", "JAX_E04_Rep", "JAX_MSK_WT")
sample2kp = which(meta_m$model_condition %in% model1)
cts_dat_m_JAX_WT = cts_dat_m[,sample2kp]
sample_info_JAX_WT = sample_info[sample2kp,]
sample_info_JAX_WT$clone = stringr::str_split_fixed(sample_info_JAX_WT$model_condition, "_", 4)[, 2]

library(edgeR)
library(limma)

# ---- inputs ----
counts  <- cts_dat_m_JAX_WT
colData <- sample_info_JAX_WT
rownames(colData) = colnames(counts)
stopifnot(all(colnames(counts) == rownames(colData)))

# ---- group factor: WT vs KO ----
group <- ifelse(colData$ko_strategy == "CTRL_KO_WT" | colData$ko_gene == "WT" | colData$model_condition == "JAX_MSK_WT",
                "WT", "KO")
group <- factor(group, levels = c("WT","KO"))

design2 <- model.matrix(~ 0 + group)
colnames(design2) <- c("WT","KO")

# ---- block vector: KO keep clone IDs; WT each unique ----
isWT  <- group == "WT"
block <- as.character(colData$clone)
block[isWT] <- paste0("WT_", seq_len(sum(isWT)))

# ---- normalization + voom (quality weights optional) ----
y2 <- DGEList(counts)
y2 <- calcNormFactors(y2)
v2 <- voom(y2, design2, plot = FALSE)

# ---- estimate correlation & fit ----
corfit2 <- duplicateCorrelation(v2, design2, block = block)
message(sprintf("Estimated intra-clone correlation: %.3f", corfit2$consensus))

fitA <- lmFit(v2, design2, block = block, correlation = corfit2$consensus)

# ---- KO vs WT contrast ----
L2 <- makeContrasts(KOvsWT = KO - WT, levels = design2)
fitA <- eBayes(contrasts.fit(fitA, L2))
ttA   <- topTable(fitA, coef = "KOvsWT", number = Inf)




model1 = c("MSK_Clone1_Rep", "MSK_Clone2_Rep", "JAX_MSK_WT")
sample2kp = which(meta_m$model_condition %in% model1)
cts_dat_m_MSK_WT = cts_dat_m[,sample2kp]
sample_info_MSK_WT = sample_info[sample2kp,]
sample_info_MSK_WT$clone = stringr::str_split_fixed(sample_info_MSK_WT$model_condition, "_", 4)[, 2]


# ---- inputs ----
counts  <- cts_dat_m_MSK_WT
colData <- sample_info_MSK_WT
rownames(colData) = colnames(counts)
stopifnot(all(colnames(counts) == rownames(colData)))

# ---- group factor: WT vs KO ----
group <- ifelse(colData$ko_strategy == "CTRL_KO_WT" | colData$ko_gene == "WT" | colData$model_condition == "JAX_MSK_WT",
                "WT", "KO")
group <- factor(group, levels = c("WT","KO"))

design2 <- model.matrix(~ 0 + group)
colnames(design2) <- c("WT","KO")

# ---- block vector: KO keep clone IDs; WT each unique ----
isWT  <- group == "WT"
block <- as.character(colData$clone)
block[isWT] <- paste0("WT_", seq_len(sum(isWT)))

# ---- normalization + voom (quality weights optional) ----
y2 <- DGEList(counts)
y2 <- calcNormFactors(y2)
v2 <- voom(y2, design2, plot = FALSE)

# ---- estimate correlation & fit ----
corfit2 <- duplicateCorrelation(v2, design2, block = block)
message(sprintf("Estimated intra-clone correlation: %.3f", corfit2$consensus))

fitB <- lmFit(v2, design2, block = block, correlation = corfit2$consensus)

# ---- KO vs WT contrast ----
L2 <- makeContrasts(KOvsWT = KO - WT, levels = design2)
fitB2 <- eBayes(contrasts.fit(fitB, L2))
ttB   <- topTable(fitB2, coef = "KOvsWT", number = Inf)

```
```{r}
hist(ttA$adj.P.Val, main = "Histogram of JAX vs. WT")
hist(ttB$adj.P.Val, main = "Histogram of MSK vs. WT")

sum(ttA$adj.P.Val > 0.5) / nrow(ttA)
sum(ttB$adj.P.Val > 0.5) / nrow(ttB)
```

```{r}
ttData = ttA[, c(1, 5)]
ttB2 = ttB
colnames(ttData) = c("logFC_JAX", "adj_p_JAX")
ttB2 = ttB2[match(rownames(ttA), rownames(ttB2)), c(1, 5)]
colnames(ttB2) = c("logFC_MSK", "adj_p_MSK")
ttData = cbind(ttData, ttB2)
ggplot(ttData, aes(-log10(adj_p_JAX), -log10(adj_p_MSK))) + geom_pointdensity() + scale_fill_gradient(trans = "log10") + 
  ggtitle("scatterplot of -log10 adj. p-values between JAX and MSK") + xlab("JAX") + ylab("MSK") + theme_classic(base_size = 15) + geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")

ggplot(ttData, aes(logFC_JAX, logFC_MSK)) + geom_pointdensity() + scale_fill_gradient(trans = "log10") + 
  ggtitle("scatterplot of logFC between JAX and MSK") + xlab("JAX") + ylab("MSK") + theme_classic(base_size = 15) + geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")
```

```{r}
# Add a significance column for coloring
ttA$threshold <- ttA$adj.P.Val < 0.05 & abs(ttA$logFC) > 1

ggplot(ttA, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = threshold), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red")) +
  theme_minimal(base_size = 14) +
  labs(title = "Volcano Plot: KO JAX vs WT",
       x = "log2 Fold Change",
       y = "-log10(p-value)") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black")

# Add a significance column for coloring
ttB$threshold <- ttB$adj.P.Val < 0.05 & abs(ttB$logFC) > 1

ggplot(ttB, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = threshold), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red")) +
  theme_minimal(base_size = 14) +
  labs(title = "Volcano Plot: KO MSK vs WT",
       x = "log2 Fold Change",
       y = "-log10(p-value)") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black")
```

```{r}
gene_anno[match(rownames(ttA), gene_anno$ensembl_gene_id), ]$hgnc_symbol[1:10]
gene_anno[match(rownames(ttB), gene_anno$ensembl_gene_id), ]$hgnc_symbol[1:10]
```

```{r}
sum(rownames(ttA[ttA$threshold, ]) %in% rownames(ttB[ttB$threshold, ])) / nrow(ttA)
sum(rownames(ttB[ttB$threshold, ]) %in% rownames(ttA[ttA$threshold, ])) / nrow(ttA)
```

```{r}
q75 = apply(cts_dat_m, 1, quantile, probs=0.75)
cts_dat_m_JAX_MSK_WT = t(cts_dat_m[q75 > 0,])
cts_dat_m_JAX_MSK_WT = log(median(meta_m$q75)*cts_dat_m_JAX_MSK_WT/meta_m$q75 + 1)
dim(cts_dat_m_JAX_MSK_WT)
meta_m$clone = stringr::str_split_fixed(meta_m$model_condition, "_", 4)[, 2]
meta_m$clone[meta_m$clone == "MSK"] = "WT"

sample_pca = prcomp(cts_dat_m_JAX_MSK_WT, center = TRUE, scale. = TRUE)
names(sample_pca)
pc_scores = sample_pca$x
eigen_vals = sample_pca$sdev^2
eigen_vals[1:5]/sum(eigen_vals)
pvs = round(eigen_vals[1:5]/sum(eigen_vals)*100,1)
pvs

PC_df = data.frame(pc_scores)
PC_df = cbind(PC_df, meta_m)

gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=clone, color=ko_strategy)) +
  geom_point(size=2.5, alpha=0.7) + scale_color_brewer(palette = "Dark2") +
  scale_shape_prism() + 
  labs(color="KO type", shape ="KO gene") + 
  xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
  ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
  ggtitle("Wild type samples vs. JAX & MSK samples") + 
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  ) +
  theme(
    legend.margin = margin(0, 0, 0, 0), 
    legend.box.just = "left", 
    legend.direction = "vertical" 
  )

print(gPC)
```

```{r, fig.height = 4.8, fig.width=15.2}
sample_info_JAX_WT$sample_names = paste0(sample_info_JAX_WT$model_condition, c(1, 1, 2, 1, 2, 3, 2, 3, 3))

counts  <- cts_dat_m_JAX_WT
colData <- sample_info_JAX_WT
rownames(colData) = colnames(counts)
stopifnot(all(colnames(counts) == rownames(colData)))

# ---- group factor: WT vs KO ----
group <- ifelse(colData$ko_strategy == "CTRL_KO_WT" | colData$ko_gene == "WT" | colData$model_condition == "JAX_MSK_WT",
                "WT", "KO")
group <- factor(group, levels = c("WT","KO"))

design2 <- model.matrix(~ 0 + group)
colnames(design2) <- c("WT","KO")

# ---- block vector: KO keep clone IDs; WT each unique ----
isWT  <- group == "WT"
block <- as.character(colData$clone)
block[isWT] <- paste0("WT_", seq_len(sum(isWT)))

# ---- normalization + voom (quality weights optional) ----
y2 <- DGEList(counts)
y2 <- calcNormFactors(y2)
# v2 <- voomWithQualityWeights(y2, design2, plot = FALSE)  # optional
v2 <- voom(y2, design2, plot = FALSE)
colnames(v2$E) = sample_info_JAX_WT$sample_names
pheatmap(t(v2$E), cluster_rows = TRUE, show_colnames = FALSE, scale = "column", fontsize = 15)

sample_info_MSK_WT$sample_names = paste0(sample_info_MSK_WT$model_condition, c(1, 1, 1, 2, 3, 2, 2, 3, 3))

counts  <- cts_dat_m_MSK_WT
colData <- sample_info_MSK_WT
rownames(colData) = colnames(counts)
stopifnot(all(colnames(counts) == rownames(colData)))

# ---- group factor: WT vs KO ----
group <- ifelse(colData$ko_strategy == "CTRL_KO_WT" | colData$ko_gene == "WT" | colData$model_condition == "JAX_MSK_WT",
                "WT", "KO")
group <- factor(group, levels = c("WT","KO"))

design2 <- model.matrix(~ 0 + group)
colnames(design2) <- c("WT","KO")

# ---- block vector: KO keep clone IDs; WT each unique ----
isWT  <- group == "WT"
block <- as.character(colData$clone)
block[isWT] <- paste0("WT_", seq_len(sum(isWT)))

# ---- normalization + voom (quality weights optional) ----
y2 <- DGEList(counts)
y2 <- calcNormFactors(y2)
# v2 <- voomWithQualityWeights(y2, design2, plot = FALSE)  # optional
v2 <- voom(y2, design2, plot = FALSE)
colnames(v2$E) = sample_info_MSK_WT$sample_names
pheatmap(t(v2$E), cluster_rows = TRUE, show_colnames = FALSE, scale = "column", fontsize = 15)

```

### DE analysis for all genes
```{r fig.height=3.2, fig.width=4.8, message=FALSE, warning=FALSE}
meta$model_condition = stringr::str_sub(meta$model_system, start = 1, end = -2)

df_size = NULL
model1 = c("JAX_C01_Rep", "JAX_E04_Rep", "MSK_Clone1_Rep", "MSK_Clone2_Rep", "JAX_MSK_WT")
sample2kp = which(meta$model_condition %in% model1)

cts_dat_m = cts_dat[,sample2kp]
meta_m    = meta[sample2kp,]

table(meta_m$ko_strategy)
stopifnot(all(meta_m$file_id == colnames(cts_dat_m)))
  
## Generate sample information matrix
cols2kp = c("model_condition", "ko_strategy", "ko_gene", 
            "run_id", "run_id_short", "timepoint", "q75")
sample_info = meta_m[,cols2kp]
colnames(sample_info)[which(colnames(sample_info)=="timepoint")] = "day"

dim(sample_info)
sample_info[1:2,]
  
# do two steps of filtering
# first, filter based on q75 of each gene
# second, filter based on whether each gene is expressed in at least 75% of the samples

dg_q75 = apply(cts_dat_m, 1, quantile, probs=0.75)
cts_dat_m = cts_dat_m[dg_q75 > 0, ]

print(sprintf("After filtering by gene expression q75, the number of genes reduces from %d to %d", 
              length(dg_q75), nrow(cts_dat_m)))

# update the q75 based on only genes that will be used in the DE analysis
sample_info$q75 = apply(cts_dat_m, 2, quantile, probs = 0.75)
sample_info$log_q75 = log(sample_info$q75)

model1 = c("JAX_C01_Rep", "JAX_E04_Rep", "JAX_MSK_WT")
sample2kp = which(meta_m$model_condition %in% model1)
cts_dat_m_JAX_WT = cts_dat_m[,sample2kp]
sample_info_JAX_WT = sample_info[sample2kp,]
sample_info_JAX_WT$clone = stringr::str_split_fixed(sample_info_JAX_WT$model_condition, "_", 4)[, 2]

library(edgeR)
library(limma)

# ---- inputs ----
counts  <- cts_dat_m_JAX_WT
colData <- sample_info_JAX_WT
rownames(colData) = colnames(counts)
stopifnot(all(colnames(counts) == rownames(colData)))

# ---- group factor: WT vs KO ----
group <- ifelse(colData$ko_strategy == "CTRL_KO_WT" | colData$ko_gene == "WT" | colData$model_condition == "JAX_MSK_WT",
                "WT", "KO")
group <- factor(group, levels = c("WT","KO"))

design2 <- model.matrix(~ 0 + group)
colnames(design2) <- c("WT","KO")

# ---- block vector: KO keep clone IDs; WT each unique ----
isWT  <- group == "WT"
block <- as.character(colData$clone)
block[isWT] <- paste0("WT_", seq_len(sum(isWT)))

# ---- normalization + voom (quality weights optional) ----
y2 <- DGEList(counts)
y2 <- calcNormFactors(y2)
v2 <- voom(y2, design2, plot = FALSE)

# ---- estimate correlation & fit ----
corfit2 <- duplicateCorrelation(v2, design2, block = block)
message(sprintf("Estimated intra-clone correlation: %.3f", corfit2$consensus))

fitA <- lmFit(v2, design2)

# ---- KO vs WT contrast ----
L2 <- makeContrasts(KOvsWT = KO - WT, levels = design2)
fitA <- eBayes(contrasts.fit(fitA, L2))
ttA   <- topTable(fitA, coef = "KOvsWT", number = Inf)




model1 = c("MSK_Clone1_Rep", "MSK_Clone2_Rep", "JAX_MSK_WT")
sample2kp = which(meta_m$model_condition %in% model1)
cts_dat_m_MSK_WT = cts_dat_m[,sample2kp]
sample_info_MSK_WT = sample_info[sample2kp,]
sample_info_MSK_WT$clone = stringr::str_split_fixed(sample_info_MSK_WT$model_condition, "_", 4)[, 2]


# ---- inputs ----
counts  <- cts_dat_m_MSK_WT
colData <- sample_info_MSK_WT
rownames(colData) = colnames(counts)
stopifnot(all(colnames(counts) == rownames(colData)))

# ---- group factor: WT vs KO ----
group <- ifelse(colData$ko_strategy == "CTRL_KO_WT" | colData$ko_gene == "WT" | colData$model_condition == "JAX_MSK_WT",
                "WT", "KO")
group <- factor(group, levels = c("WT","KO"))

design2 <- model.matrix(~ 0 + group)
colnames(design2) <- c("WT","KO")

# ---- block vector: KO keep clone IDs; WT each unique ----
isWT  <- group == "WT"
block <- as.character(colData$clone)
block[isWT] <- paste0("WT_", seq_len(sum(isWT)))

# ---- normalization + voom (quality weights optional) ----
y2 <- DGEList(counts)
y2 <- calcNormFactors(y2)
# v2 <- voomWithQualityWeights(y2, design2, plot = FALSE)  # optional
v2 <- voom(y2, design2, plot = FALSE)

# ---- estimate correlation & fit ----
corfit2 <- duplicateCorrelation(v2, design2, block = block)
message(sprintf("Estimated intra-clone correlation: %.3f", corfit2$consensus))

fitB <- lmFit(v2, design2)

# ---- KO vs WT contrast ----
L2 <- makeContrasts(KOvsWT = KO - WT, levels = design2)
fitB2 <- eBayes(contrasts.fit(fitB, L2))
ttB   <- topTable(fitB2, coef = "KOvsWT", number = Inf)

```

```{r}
# Add a significance column for coloring
ttA$threshold <- ttA$adj.P.Val < 0.05 & abs(ttA$logFC) > 1

ggplot(ttA, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = threshold), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red")) +
  theme_minimal(base_size = 14) +
  labs(title = "Volcano Plot: KO JAX vs WT without clone info",
       x = "log2 Fold Change",
       y = "-log10(p-value)") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black")

# Add a significance column for coloring
ttB$threshold <- ttB$adj.P.Val < 0.05 & abs(ttB$logFC) > 1

ggplot(ttB, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = threshold), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red")) +
  theme_minimal(base_size = 14) +
  labs(title = "Volcano Plot: KO MSK vs WT without clone info",
       x = "log2 Fold Change",
       y = "-log10(p-value)") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black")
```

```{r}
ttDataNoClone = ttA[, c(1, 5)]
ttDataWithClone = ttData
ttB2 = ttB
colnames(ttDataNoClone) = c("logFC_JAX", "adj_p_JAX")
ttB2 = ttB2[match(rownames(ttA), rownames(ttB2)), c(1, 5)]
colnames(ttB2) = c("logFC_MSK", "adj_p_MSK")
ttDataNoClone = cbind(ttDataNoClone, ttB2)
ggplot(ttDataNoClone, aes(-log10(adj_p_JAX), -log10(adj_p_MSK))) + geom_pointdensity() + scale_fill_gradient(trans = "log10") + 
  ggtitle("scatterplot of adj. p-values between JAX and MSK") + xlab("JAX") + ylab("MSK")+ theme_classic(base_size = 15) + geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")

ggplot(ttDataNoClone, aes(logFC_JAX, logFC_MSK)) + geom_pointdensity() + scale_fill_gradient(trans = "log10") +  
  ggtitle("scatterplot of logFC between JAX and MSK") + xlab("JAX") + ylab("MSK")+ theme_classic(base_size = 15) + geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")

colnames(ttDataNoClone) = paste0(colnames(ttDataNoClone), "NoClone")
ttDataNoClone = ttDataNoClone[match(rownames(ttDataWithClone), rownames(ttDataNoClone)), ]
ttDataWithClone = cbind(ttDataWithClone, ttDataNoClone)

ggplot(ttDataWithClone, aes(-log10(adj_p_JAX), -log10(adj_p_JAXNoClone))) + geom_pointdensity() + scale_fill_gradient(trans = "log10") + 
  ggtitle("scatterplot of -log10 adj. p-values between JAX and JAX with no clone") + xlab("JAX") + ylab("JAX with no clone")+ theme_classic(base_size = 12) + geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")

```

```{r}
set.seed(11)
ngenes   <- 1000
nsamples <- 9
grp      <- c(0,0,0,0,0,0,1,1,1)           # Grp2 vs Grp1 indicator
design   <- cbind(Intercept=1, Grp2vs1=grp)
block    <- c(1,1,1, 2,2,2, 3,4,5)         # first 6 samples in 2 blocks of size 3; last 3 singletons
rho <- 0.30   # desired intra-block correlation
# --------------------------------------------------------
# choose many DE genes (positive and negative) + effect sizes
# --------------------------------------------------------
prop_up   <- 0.15
prop_down <- 0.15
n_up      <- round(prop_up   * ngenes)
n_down    <- round(prop_down * ngenes)
n_null    <- ngenes - n_up - n_down

# Effect size distribution (you can tune these)
# positive effects
beta_up   <- rnorm(n_up,   mean = 0.8, sd = 0.25)
# negative effects
beta_down <- rnorm(n_down, mean = -0.8, sd = 0.25)
# null effects
beta_null <- rep(0, n_null)
beta <- c(beta_up, beta_down, beta_null)
beta <- beta[sample.int(ngenes)]  # shuffle across genes

# Precompute block-level effects per gene:
# Var(block effect) : Var(residual) ratio chosen to yield correlation ~ rho
# If we set Var(block) = rho and Var(resid)=1-rho, correlation ≈ rho (before sigma2_g scaling)
n_blocks <- length(unique(block))
block_index <- match(block, sort(unique(block)))
y <- matrix(0, nrow=ngenes, ncol=nsamples)

for (g in 1:ngenes) {
  # block random effects (shared within block for this gene)
  b_eff <- rnorm(n_blocks)
  # residual noise
  eps   <- rnorm(nsamples)
  # group effect
  mu    <- beta[g] * grp
  # assemble
  y[g, ] <- b_eff[block_index] + eps + mu * 0.5
  y[g, 1:6] = y[g, 1:6] + beta[g] * 3
}

rownames(y) <- paste0("Gene", seq_len(ngenes))
colnames(y) <- paste0("S", seq_len(nsamples))
pheatmap::pheatmap(y, scale = "row", show_rownames = FALSE, cluster_cols = FALSE)
# -------------------------
# estimate correlation & fit
# -------------------------
dupcor <- duplicateCorrelation(y, design, block=block)
cat("Estimated consensus correlation:", round(dupcor$consensus.correlation, 3), "\n")

fit_cor <- lmFit(y, design, block=block, correlation=dupcor$consensus)
fit_cor <- eBayes(fit_cor)

fit_iid <- lmFit(y, design)
fit_iid <- eBayes(fit_iid)

tt_cor <- topTable(fit_cor, coef="Grp2vs1", number=Inf)
tt_iid <- topTable(fit_iid, coef="Grp2vs1", number=Inf)
tt_iid <- tt_iid[match(rownames(tt_cor), rownames(tt_iid)), ]

# quick checks
# how many DE genes did we plant?
table_planted <- table(cut(beta, c(-Inf, -1e-12, 1e-12, Inf), labels=c("down","null","up")))
print(table_planted)

# how many recovered at FDR 5% (with correlation model)?
fdr <- 0.05
sig_cor <- sum(tt_cor$adj.P.Val < fdr)
sig_iid <- sum(tt_iid$adj.P.Val < fdr)
cat("Significant (FDR 5%) — with corr:", sig_cor, "without corr:", sig_iid, "\n")

# optional: compare adj p-values under the two fits
data.frame(WCor = -log10(tt_cor$adj.P.Val), OCor = -log10(tt_iid$adj.P.Val)) %>% ggplot(aes(WCor, OCor)) + geom_pointdensity() + scale_fill_gradient(trans = "log10") + 
  ggtitle("Simulated scatterplot of -log10 adj.p-val with and without DupCor") + xlab("With DupCor") + ylab("Without DupCor")+ theme_classic(base_size = 10)

```

```{r}
gc()
sessionInfo()
```
